Tidy data structure to support exploration and modeling of temporal-context data

### 2018-04-24 main.Rmd
started to mention time series and panel data

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntatical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a transparent and human-readable workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The tidy temporal data along with its supporting methods has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \cdots, N$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping variables exhibiting nested or crossed structures, linking to other data tables, and etc.

Although a matrix is a tabular form, it assumes homogeneity (that is, all the columns must be real values.) and time indices implicitly inferred as attributes (or meta-information). This format of data input is model-oriented rather than data-oriented as it proves computationally efficient for many statistical models when doing matrix operations. However, it is has the effect of discarding interesting information in the data, is more opaque in the handling of the temporal components, and less efficient for many other data analysis tasks, such as data transformation, visualisation and model diagnostics, like forcing the fitting of a square peg into a round hole.

In addition, time indices form the basis of temporal data and the source of contextual information to understand temporal dependence. Time as attributes limits the use of time context, since it is difficult to access and manipulate. For example, the components of date-times, such as months, day of week, and time of day, cannot be easily extracted in order to examine seasonal fluctuations. It is also impossible to switch between different time zones and daylight savings to compare time across places when needed.

This paper proposes a tidy data structure and a modern data pipeline for storing, managing and analysing time series data, using a collection of fluent and fluid tools to help with exploitation in temporal context.

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntatical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a transparent and human-readable workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The architecture of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \cdots, N$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

However, such richness in temporal data cannot be easily accommodated by currently available data structures in any statistical computing package. In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to column types. A matrix is used to represent multivariate time series, assuming homogeneity (that is, all the columns must be of same type.) and time indices implicitly inferred as attributes (or meta-information).

This paper proposes a tidy and unified data structure and a modern data pipeline for storing, managing and analysing time series data, using a collection of fluent and fluid tools to help with exploitation in temporal context.
